---
title: "Image Processing:  Additional Analyses"
author: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r seed}
set.seed(41)
```

```{r libraries}
library(data.table)
library(DT)
library(here)   # needed to import Part 1
```

```{r constants}

```

```{r functions}
round.numerics <- function(x, digits){
  if(is.numeric(x)){
    x <- round(x = x, digits = digits)
  }
  return(x)
}
```

```{r load_data}
# Import Part 1 so we can reuse dt_all, dev_sets, test_raw, px_cols, etc.
knitr::knit_child(here::here("image_processing.Rmd"), quiet = TRUE)

# Sanity: required objects should exist after sourcing Part 1
stopifnot(exists("dt_all"), exists("dev_sets"), exists("test_raw"), exists("px_cols"))

# Pick the best single run (lowest Points)
best_row  <- dt_all[order(Points)][1]
best_n    <- best_row$SampleSize
best_iter <- best_row$Iter
best_model<- best_row$Model

# Recreate the exact development sample used by that best run
dat_best   <- data.table::copy(dev_sets[[as.character(best_n)]][[best_iter]])
test_local <- data.table::copy(test_raw)
```

```{r explore_data, eval = FALSE}
# Optional quick peek:
# head(dt_all)
# best_row
```

```{r clean_data}
# Align factor levels
dat_best[,   label := factor(label)]
test_local[, label := factor(label, levels = levels(dat_best$label))]
```

# {.tabset}

## Introduction

## Predictive Accuracy by Product

This addendum analyzes **predictive accuracy by product category** for the **single best model** found in Part 1 (lowest overall *Points*).
Using the exact development subset that produced the winning score, we refit that model, predict on the fixed test set, and compute **per-class accuracy** (share of test images correctly classified within each label).
These class-level results reveal which product types are easiest/hardest to recognize under the current 49-pixel representation and baseline hyperparameters.

```{r per_product_accuracy}
# Lightweight switcher for the top contenders in Part 1 (RF / XGBoost / SVM).
predict_best <- function(model_name, train_dt, test_dt){
  Xtr <- as.matrix(train_dt[, ..px_cols])
  Xte <- as.matrix(test_dt[,  ..px_cols])

  if (grepl("randomForest", model_name, ignore.case = TRUE)) {
    fit <- randomForest::randomForest(
      x = train_dt[, ..px_cols],
      y = train_dt$label,
      ntree = 120, mtry = 7
    )
    predict(fit, newdata = test_dt[, ..px_cols])

  } else if (grepl("xgboost", model_name, ignore.case = TRUE)) {
    lv <- levels(train_dt$label)
    y  <- as.integer(train_dt$label) - 1L
    dtr <- xgboost::xgb.DMatrix(Xtr, label = y)
    dte <- xgboost::xgb.DMatrix(Xte)
    fit <- xgboost::xgb.train(
      params = list(objective="multi:softprob", num_class=length(lv),
                    eta=0.1, max_depth=4, subsample=0.8, colsample_bytree=0.8),
      data = dtr, nrounds = 80, verbose = 0
    )
    pr  <- predict(fit, dte)
    factor(lv[max.col(matrix(pr, nrow = nrow(test_dt), byrow = TRUE))], levels = lv)

  } else if (grepl("svm", model_name, ignore.case = TRUE)) {
    fit <- e1071::svm(
      x = as.data.frame(train_dt[, ..px_cols]),
      y = train_dt$label,
      kernel = "radial", cost = 1, gamma = 1/49, scale = TRUE
    )
    predict(fit, newdata = as.data.frame(test_dt[, ..px_cols]))

  } else {
    stop("Model handler not implemented for: ", model_name)
  }
}

# Predict with the winning model
pred <- predict_best(best_model, dat_best, test_local)

# Per-product accuracy table
acc_by_product <- data.table(label = test_local$label, pred = pred)[
  , .(n = .N, accuracy = mean(label == pred)), by = label][
  order(-accuracy)]

acc_by_product[, accuracy := round(accuracy, 4)]

DT::datatable(
  acc_by_product,
  rownames = FALSE,
  options = list(pageLength = 10),
  caption = "Predictive Accuracy by Product (best model, exact best-run dev subset)"
)
```

### Interpretation

Performance varies meaningfully by class. Categories with **distinct silhouettes/textures** (e.g., boots or sneakers) tend to score higher accuracy, while **look-alike apparel** (e.g., shirt vs. pullover) is more confusable in a 7×7 pixel grid.
To improve weaker classes, consider (i) **feature learning** (e.g., CNNs or embeddings), (ii) **targeted augmentation** to diversify edge cases, and (iii) **class-aware tuning** or cost-sensitive training so that rare/confusable categories receive proportionally more modeling attention.

## Independent Investigation


