---
title: "Data Science Consulting:  Midterm Team Project -- Part 1"
author: "Clara Dragut, Yuwen Wang, Ziyao Wang"
date: "Nov 2025"
output: html_document
---

```{r setup, include=FALSE}
set.seed(72)
knitr::opts_chunk$set(echo = TRUE, comment="", warning = FALSE, message = FALSE, tidy.opts=list(width.cutoff=55))
```

```{r libraries, echo = FALSE}
library(data.table)
library(here)
library(knitr)
library(rmarkdown)

# ML models
library(randomForest)
```

```{r source_files}

```

```{r functions}



```

```{r load_data}
train_path <- here::here("data", "MNIST-fashion training set-49.csv")
test_path  <- here::here("data", "MNIST-fashion testing set-49.csv")

train_raw <- data.table::fread(train_path)
test_raw  <- data.table::fread(test_path)
```

```{r clean_data}

```

```{r generate_samples}
# pick an index to visualize
i <- 1

# extract pixel values (drop the text label)
px <- as.numeric(train_raw[i, -1])

# convert to 7×7 matrix
mat <- matrix(px, nrow = 7, ncol = 7, byrow = TRUE)

# plot
image(t(apply(mat, 2, rev)), col = gray.colors(256),
      main = train_raw$label[i], axes = FALSE)
```

## Introduction {.tabset}


### Model 1


```{r code_model1_development, eval = TRUE}

```

```{r load_model1}

```

### Model 2


```{r code_model2_development, eval = TRUE}

```

```{r load_model2}

```

### Model 3


```{r code_model3_development, eval = TRUE}

```

```{r load_model3}

```


### Model 4 - Random Forest
```{r code_model4_development, eval = TRUE}
# pixel columns
px_cols <- paste0("pixel", 1:49)

# ensure label is factor
train_raw[, label := as.factor(label)]
test_raw[,  label := factor(label, levels = levels(train_raw$label))]

# record time (train + predict)
t0 <- proc.time()

# light hyperparameter tuning (keep simple)
# try a slightly larger mtry and slightly fewer trees to balance accuracy + time
best_ntree <- 150
best_mtry  <- 9

```

```{r load_model4}
# fit model
rf_model <- randomForest(
  x = train_raw[, ..px_cols],
  y = train_raw$label,
  ntree = best_ntree,
  mtry  = best_mtry
)

# predict
rf_pred <- predict(rf_model, newdata = test_raw[, ..px_cols])

# total elapsed seconds
time_sec <- (proc.time() - t0)[["elapsed"]]

# scoring components
A <- nrow(train_raw) / 60000
B <- min(1, time_sec / 60)
C <- mean(rf_pred != test_raw$label)
Points <- 0.15*A + 0.10*B + 0.75*C

# show results
list(
  ntree = best_ntree,
  mtry  = best_mtry,
  A = A,
  B = B,
  C = C,
  Points = Points,
  time_sec = time_sec
)
```


### Model 5 - Elastic Net

```{r code_model5_development, eval = TRUE}
library(glmnet)
library(data.table)

# pixel columns
px_cols <- paste0("pixel", 1:49)

# ensure label is factor
train_raw[, label := as.factor(label)]
test_raw[,  label := factor(label, levels = levels(train_raw$label))]

# convert to model matrices for glmnet
x_train <- as.matrix(train_raw[, ..px_cols])
y_train <- train_raw$label
x_test  <- as.matrix(test_raw[, ..px_cols])
y_test  <- test_raw$label

# record time (train + predict)
t0 <- proc.time()

# light hyperparameter tuning
alpha_val <- 0.3     # elastic net mixing
```

```{r load_model5}
# fit model
fit <- glmnet(
  x = x_train,
  y = y_train,
  alpha = alpha_val,
  family = "multinomial"
)

# predict class labels
preds <- predict(fit, newx = x_test, s = fit$lambda.min, type = "class")

# total elapsed seconds
time_sec <- (proc.time() - t0)[["elapsed"]]

# scoring components
A <- nrow(train_raw) / 60000
B <- min(1, time_sec / 60)
C <- mean(preds != y_test)
Points <- 0.15*A + 0.10*B + 0.75*C

# show results
list(
  alpha = alpha_val,
  A = A,
  B = B,
  C = C,
  Points = Points,
  time_sec = time_sec
)
```

### Model 6


```{r code_model6_development, eval = TRUE}

```

```{r load_model6}

```

### Model 7


```{r code_model7_development, eval = TRUE}

```

```{r load_model7}

```

### Model 8


```{r code_model8_development, eval = TRUE}
library(gbm)

px_cols <- paste0("pixel", 1:49)

train_raw[, label := as.factor(label)]
test_raw[,  label := factor(label, levels = levels(train_raw$label))]

lbl_levels <- levels(train_raw$label)
y_train_num <- as.integer(train_raw$label) - 1L

gbm_params <- list(
  n.trees = 500,
  interaction.depth = 3,
  shrinkage = 0.05,
  n.minobsinnode = 10,
  bag.fraction = 0.8
)

t0 <- proc.time()

```

```{r load_model8}
gbm_fit <- gbm::gbm(
  formula = y_train_num ~ .,
  data    = data.frame(y_train_num = y_train_num, train_raw[, ..px_cols]),
  distribution = "multinomial",
  n.trees = gbm_params$n.trees,
  interaction.depth = gbm_params$interaction.depth,
  shrinkage = gbm_params$shrinkage,
  n.minobsinnode = gbm_params$n.minobsinnode,
  bag.fraction = gbm_params$bag.fraction,
  train.fraction = 1.0,
  verbose = FALSE
)

# predict probs -> class
pred_prob <- predict(
  gbm_fit,
  newdata = as.data.frame(test_raw[, ..px_cols]),
  n.trees = gbm_params$n.trees,
  type = "response"
)
prob_mat <- matrix(pred_prob, nrow = nrow(test_raw), ncol = length(lbl_levels))
pred_idx <- max.col(prob_mat, ties.method = "first")
pred_m8 <- factor(lbl_levels[pred_idx], levels = lbl_levels)

time_sec <- (proc.time() - t0)[["elapsed"]]
A <- nrow(train_raw) / 60000
B <- min(1, time_sec / 60)
C <- mean(pred_m8 != test_raw$label)
Points <- 0.15*A + 0.10*B + 0.75*C

res_m8 <- list(Model = "M8_gbm", A=A, B=B, C=C, Points=Points, time_sec=time_sec)

res_m8
```

### Model 9


```{r code_model9_development, eval = TRUE}
px_cols <- paste0("pixel", 1:49)

train_raw[, label := as.factor(label)]
test_raw[,  label := factor(label, levels = levels(train_raw$label))]

x_train_df <- as.data.frame(train_raw[, ..px_cols])
x_test_df  <- as.data.frame(test_raw[, ..px_cols])
y_train    <- train_raw$label
y_test     <- test_raw$label

svm_cost  <- 1
svm_gamma <- 1/49

t0 <- proc.time()
```

```{r load_model9}
install.packages("e1071")

svm_mod <- e1071::svm(
  x = x_train_df, y = y_train,
  kernel = "radial",
  cost = svm_cost,
  gamma = svm_gamma,
  scale = TRUE
)
pred_m9 <- predict(svm_mod, newdata = x_test_df)

time_sec <- (proc.time() - t0)[["elapsed"]]
A <- nrow(train_raw) / 60000
B <- min(1, time_sec / 60)
C <- mean(pred_m9 != y_test)
Points <- 0.15*A + 0.10*B + 0.75*C

res_m9 <- list(Model = "M9_svm_rbf", A=A, B=B, C=C, Points=Points, time_sec=time_sec)

res_m9
```

### Model 10


```{r code_model10_development, eval = TRUE}
px_cols <- paste0("pixel", 1:49)

train_raw[, label := as.factor(label)]
test_raw[,  label := factor(label, levels = levels(train_raw$label))]

x_train_df <- as.data.frame(train_raw[, ..px_cols])
x_test_df  <- as.data.frame(test_raw[, ..px_cols])
y_train    <- train_raw$label
y_test     <- test_raw$label

t0 <- proc.time()
```

```{r load_model10}
nb_mod <- e1071::naiveBayes(x = x_train_df, y = y_train)
pred_m10 <- predict(nb_mod, newdata = x_test_df)

time_sec <- (proc.time() - t0)[["elapsed"]]
A <- nrow(train_raw) / 60000
B <- min(1, time_sec / 60)
C <- mean(pred_m10 != y_test)
Points <- 0.15*A + 0.10*B + 0.75*C

res_m10 <- list(Model = "M10_naive_bayes", A=A, B=B, C=C, Points=Points, time_sec=time_sec)

res_m10
```

## Scoreboard

```{r scoreboard}
library(data.table)

res_names <- ls(pattern = "^res_m\\d+$")
res_list <- lapply(res_names, function(n) {
  if (exists(n)) get(n) else NULL
})
results <- Filter(Negate(is.null), res_list)

if (length(results) == 0) {
  cat("No model summaries found. Knit the model chunks first.")
} else {
  dt <- rbindlist(lapply(results, as.data.table), fill = TRUE)
  setorder(dt, Points)
  knitr::kable(dt, digits = 4, caption = "Scoreboard")
}
```

## Discussion


## Model Development Responsibilities

For the 10 models, please list the names of the developers along with percentages for how the responsibilities were divided.

1.  
2. 
3. 
4. Yuwen Wang
5. Yuwen Wang
6.
7.
8. Clara Dragut
9. Clara Dragut
10. Clara Dragut

## References


